---
title: "Final"
output: html_document
---
```{r}
setwd("/Users/jessicasaini/Desktop/UoW/Stat 847/Final Project")
library(GGally)
library(tidyverse)
library(mice)
library(sentimentr)
library(plyr)
```

#Reading The data
```{r}
df = read.csv("Gamelog T20I Stat 847.csv")
head(df)
#View(df)

df2 = df%>% filter(df$Inning==2) 
text<- df2$FullNotes
head(text)
```
#Question 1

```{r}
t20_df = subset(df, df$Format=="T20I")
#View(t20_df)

#Check for NA values

#Change -1 in NumOutcome to 0 
t20_df$NumOutcome[t20_df$NumOutcome == -1]<-0
t20_df$TotalRunsAtThisInstant =0
batsmen = unique(t20_df$Batsman)
```


```{r}
#matches = unique(t20_df$MatchNo)
matches = c(33,161)
count =0 
for(i in matches){
  runs = 0
  dataset <- subset(t20_df,MatchNo %in% i)
  for(j in 1:nrow(dataset)){
    #Loop over every row
    row = dataset[j,]
    runs = runs + row$NumOutcome
    #print(runs)
    row$TotalRunsAtThisInstant = runs
    dataset[j,22] = runs
    t20_df[t20_df$MatchNo==i,]
    #print(row$TotalRunsAtThisInstant)
  }
}

View(appended)
```

#Probability of a ball being "wide","leg bye","bye","wides", "leg byes", "byes"
t20_df_BallType = t20_df %>% filter(BallType %in% c("wide","leg bye","bye","wides", "leg byes", "byes"))
t20_df_BallType = t20_df %>% filter(BallType %in% c("wide","leg bye","bye","wides", "leg byes", "byes"))

ggplot(t20_df_BallType, mapping = aes(x = Ball, y = Wickets, fill = NumOutcome )) + geom_point()+
+     ylim(0,10) 

# Sentiment Analysis
```{r}
#Split your text data into sentences 
text<- get_sentences(text)

```

```{r}
analysis = sentiment(text)
```
```{r}
summary(analysis$sentiment)
analysis %>% filter(between(sentiment,-1,1)) ->  bounded_sentences
```

```{r}
View(bounded_sentences)
```


##Question 3 
Aggregate the matches (hint: the ddply
function in the K-means clustering section is great for this) and apply K-means clustering to the matches to find some meaningful archetypes of games.

Example clusters: high scoring games, very close games, games
with early wickets, etc.
```{r}

df3 = df 

df_profiles = ddply(df3, "MatchNo", summarize,
    total_runs = sum( pmax(NumOutcome, 0)),
    runs_1stinning = sum(pmax(NumOutcome*1(inning == 1), 0)),
total_wickets = length(which(NumOutcome == 1)),
fielder_mentions = length(which(Fielder != "")),
balls_until_1st_wicket = length(which(wickets == 0)),
average_wickets_in_during_match = mean(wickets, na.rm=TRUE))

```


#Testing
```{r}
for(i in matches){
runs =0 
t2<-t2%>% 
  mutate(TotalRunsInstant = if_else(t2$MatchNo==i,runs+t2$NumOutcome,0))
runs = runs+t2$NumOutcome
}
```

```{r}
matches = c(33)
for (i in matches){
  matchnum =i 
  print(matchnum)
  runs = 0
  
  for(j in 1:nrow(t2)){
  if(t2$MatchNo==matchnum){#For every match
  
    }
    
  }
  
}

```

#Relative frequency 
```{r}

a = a[, c("MatchNo", "NumOutcome", "Wickets", "Over")]
a["Rel_Freq"]<-0
freq = table(a$NumOutcome)/length(a$NumOutcome)
freq = data.frame(freq)
outcomes <- unique(a$NumOutcome)

for (i in outcomes){
for(j in 1:nrow(a)){
  if(a[j,2]==i){
    a[j,5]=freq[freq$Var1==i]
  }
}
}
View(a)
```

```{r}

a<- t20_df
a = a[, c("MatchNo", "NumOutcome", "Wickets", "Over")]
b <-a[, c("MatchNo", "NumOutcome")]
a<-a %>%
  group_by(MatchNo, NumOutcome,Wickets,Over) %>%
  dplyr::summarise(n = n())%>%
mutate(freq = n / sum(n))
View(a)
ggplot(a, mapping = aes(x = Over, y = Wickets, fill = NumOutcome )) + geom_tile()+ylim(0,10)+ scale_fill_gradientn(colors = hcl.colors(20, "RdYlGn"))
```

```{r}
#Calculation of Relative Frequency for each run
a<- t20_df
a = a[, c("MatchNo", "NumOutcome", "Wickets", "Over")]
b <-a[, c("MatchNo", "NumOutcome")]
a<-a %>%
  group_by(MatchNo, NumOutcome,Wickets,Over) %>%
  dplyr::summarise(n = n())%>%
mutate(freq = n / sum(n))
View(a)

ggplot(a, mapping = aes(x = Over, y = Wickets, fill = n )) + geom_tile()+ylim(0,10)+ scale_fill_gradientn(colors = hcl.colors(20, "RdYlGn"))

#A graph with some sense
a<-a %>%
     group_by(NumOutcome,Wickets,Over) %>%
     dplyr::summarise(n = n())%>%
     mutate(freq = n / sum(n))
ggplot(a, mapping = aes(x = Over, y = Wickets, fill = n )) + geom_tile()+ylim(0,10)+ scale_fill_gradientn(colors = hcl.colors(20, "RdYlGn"))
#Outcome of every Ball scored in T20 Innings 
ggplot(a, mapping = aes(x = Over, y = Wickets, fill = NumOutcome )) + geom_tile()+ylim(0,10)+ scale_fill_gradientn(colors = hcl.colors(20, "RdYlGn"))
```

```{r}
#Exceptional Runs Final
a<- df
a$NumOutcome[a$NumOutcome == -1]<-0
a <- subset(a,NumOutcome %in% c(3,5,7))
a<-a %>%
     group_by(MatchNo,NumOutcome,Wickets,Over) %>%
     dplyr::summarise(n = n())%>%
     mutate(freq = n / sum(n))
ggplot(a, mapping = aes(x = Over, y = Wickets, fill = NumOutcome )) + geom_tile()+ylim(0,10)+ scale_fill_gradientn(colors = hcl.colors(20, "RdYlGn"))

```

```{r}
#Bubble Chart Final
a<- df
a$NumOutcome[a$NumOutcome == -1]<-0
a <- subset(a,NumOutcome %in% c(3,5,7))
a<-a %>%
     group_by(MatchNo,NumOutcome,Wickets,Over,Inning) %>%
     dplyr::summarise(n = n())%>%
     mutate(freq = n / sum(n))

ggplot(a, aes(x=Over, y=Wickets, size = Inning,color=NumOutcome)) +
    geom_point(alpha=0.7)
```




#Question 3
#Kmeans

**Scaling of Variables:**


Graph is left

```{r}

#Convert categorical variables to numeric
gamelog = df

matches <- unique(gamelog$MatchNo)

```


```{r}
df_matches = ddply( gamelog, "MatchNo" , summarize , 
total_runs = sum( pmax(NumOutcome, 0)) , 
#total_wickets = length(which(NumOutcome == 1)), 

fielder_mentions = length(which(Fielder != "")),
balls_until_1st_wicket = length(which(Wickets == 0)) , 
average_wickets_in_during_match = mean(Wickets, na.rm=TRUE), 
format=Format[1], 
teamBowling = TeamBowling[1],
teamBatting = TeamBowling[1] 
)
```

#Fielder Mentions
```{r}
df_matches = ddply( gamelog, "MatchNo" , summarize , 
total_runs = sum( pmax(NumOutcome, 0)) , 
#total_wickets = length(which(NumOutcome == 1)), 

fielder_mentions = length(which(Fielder != "")),
balls_until_1st_wicket = length(which(Wickets == 0)) , 
average_wickets_in_during_match = mean(Wickets, na.rm=TRUE), 
format=Format[1], 
teamBowling = TeamBowling[1],
teamBatting = TeamBowling[1] 
)

```


#Games with Early Wickets, High scoring, Average Wickets and Fielder Mentions
```{r}


df_kmeans = subset(df_matches, select = c(total_runs, average_wickets_in_during_match,balls_until_1st_wicket,fielder_mentions
                                      ))

df_kmeans = na.omit(df_kmeans)
wssd <- rep(NA,9)
for(k in 2:10)
 {
     emo_clust <- kmeans(df_kmeans, centers = k)
     wssd[k-1] <- emo_clust$tot.withinss
}

 centers <- 2:10
 dat <- data.frame(centers, wssd)
 gr3 <- ggplot(dat, aes(x=centers, y=wssd)) +
     geom_line() + 
     geom_point() +
     xlab("number of clusters") +
     ylab("WSSD")
 plot(gr3)
 
#Choosing number of clusters as 4

emo_clust_4 <- kmeans(df_kmeans, centers = 4)
emo_clust_4$centers

msd <- sqrt(emo_clust_4$withinss / emo_clust_4$size)
print(msd)

print(emo_clust_4$size)


#Plot the graphs
#ggplot(df_matches, mapping=aes(x=total_runs,y=total_wickets)) + geom_density_2d_filled()




```


#Games with Early Wickets, High scoring, Average Wickets
```{r}
runs_1stinning = sum(pmax(numOutcome*1(inning == 1), 0)))

df_kmeans = subset(df_matches, select = c(total_runs, average_wickets_in_during_match,balls_until_1st_wicket
                                      ))

df_kmeans = na.omit(df_kmeans)
wssd <- rep(NA,9)
for(k in 2:10)
 {
     emo_clust <- kmeans(df_kmeans, centers = k)
     wssd[k-1] <- emo_clust$tot.withinss
}

 centers <- 2:10
 dat <- data.frame(centers, wssd)
 gr3 <- ggplot(dat, aes(x=centers, y=wssd)) +
     geom_line() + 
     geom_point() +
     xlab("number of clusters") +
     ylab("WSSD")
 plot(gr3)
 
#Choosing number of clusters as 4

emo_clust_4 <- kmeans(df_kmeans, centers = 4)
emo_clust_4$centers

msd <- sqrt(emo_clust_4$withinss / emo_clust_4$size)
print(msd)

print(emo_clust_4$size)


#Plot the graphs
ggplot(df_matches, mapping=aes(x=total_runs,y=total_wickets)) + geom_density_2d_filled()

